Understanding meta-analysis through data simulation with applications to power analysis

### 2023-06-29 documents/paper/paper.Rmd
fixing fixed vs random-effects figure caption

```{r monte-carlo-simulation}
set.seed(123) # seed for simulation to ensure the reproducibility of results
theta <- 0.5 # real mean difference in the populations
sigma <- 1 # real standard deviations in the population
n <- 30 # sample size for both groups
nsims <- 1e5 # number of simulations = 100000

# simulates the sampling distribution of the mean difference
d_mc <- replicate(nsims, expr = {
  g1 <- rnorm(n = n, mean = theta, sd = sigma)
  g2 <- rnorm(n = n, mean = 0, sd = sigma)
  mean(g1) - mean(g2) # calculates the sample mean difference and returns the value
})

@@

```{r notation-tab, echo=FALSE, result = "asis"}
notation_table <- tibble::tribble(
     ~`Equation`,     ~`Code`,                                               ~`Description`,
       "$\\theta$",   "theta",                  "Fixed-effects model real effect size",
       "$\\mu$",      "mu",                        "Random-effects model real effect size",
        "$\\tau^2$",    "tau2",                            "Effect size heterogeneity",
       "$\\tau^2_r$",   "tau2r",                   "Residual effect size heterogeneity",
  "$\\bar y_{T, C}$",        "",               "Mean of the experimental/control group",
      "$s_{T, C}$",        "", "Standard deviation of the experimental/control group",
        "$n_{T, C}$",  "nt, nc",        "Sample size of the experimental/control group",
             "$d_i$",      "yi",                                 "Observed effect size",
      "$$\\sigma_{\\epsilon_i}^{2}$$",      "vi",                          "Observed sampling variance",
       "$\\theta_i$", "theta\\_i",              "Random effect for the study i",
        "$\\beta_0$",      "b0",                            "Meta-regression intercept",
         "$\\beta_1$",      "b1",                                "Meta-regression slope",
               "$k$",       "k",                                    "Number of studies",
     "$\\epsilon_i$",        "",             "Sampling error for the study i"
  )

@@

```{r sim-random-effect}
set.seed(seed)
k <- 30 # number of studies
n <- 20 # number of participants per group, per study
mu_theta <- 0.3 # real effect size
tau2 <- 0.2 # the effect size heterogeneity

sim <- make_data(k = k, nc = n, nt = n, mu_theta = mu_theta)

@@

# adding the by-study adjustment to the average effect
sim$es <- sim$mu_theta + sim$theta_i

@@

```{r random-effect-high-low-precision, echo = FALSE, fig.asp=0.8, fig.cap=get_caption()}
set.seed(seed)
k <- 10 # number of studies
mu_theta <- 0.3 # real effect size
tau2 <- 0.2
n <- c(30, 500)

sim <- make_data(k = k, mu_theta = mu_theta)

@@

sim$es <- sim$mu_theta + sim$theta_i
sim <- sim_studies(sim$es, sim$nc, sim$nt, data = sim)

@@

```{r sim-random-effect-i2}
set.seed(seed)
k <- 30 # number of studies
mu_theta <- 0.3 # real effect size
n <- 30 # sample size per group, per study
I2 <- 0.6 # desired I2 value

v <- (n + n)/(n * n) + mu_theta^2/(2 * (n + n - 2)) # typical within study variance
tau2 <- -((I2*v)/(I2 - 1))

sim <- make_data(k = k, nc = n, nt = n, mu_theta = mu_theta)

sim$theta_i <- rnorm(k, 0, sqrt(tau2))
sim$es <- sim$mu_theta + sim$theta_i 
sim <- sim_studies(sim$es, sim$nc, sim$nt, data = sim)
res <- rma(yi, vi, method = "REML", data = sim)
```

@@

A common example of meta-regression is by including a categorical predictor, including information about study-level features. In our example, a group of studies uses an online memory task while others use a standard lab-based task. The Equation \@ref(eq:random-effect-model1) can be easily extended for a meta-regression model by including a variable encoding the type of task (online vs. lab-based) and the expected difference between the two levels of the moderator (i.e., the *lab vs. online effect*). In regression terms, we could use a dummy variable ($X_{1}$) that takes the value of 0 for the *lab-based task* (*l*) and a value of 1 for an *online task* (*o*)^[The *dummy coding* (also known as treatment coding) is the default in R and `metafor` but other coding schemes could be used [see @Schad2020-ht for an overview of contrast coding schemes]]. Now we fix $\beta_{1}$ to be the *lab vs. online effect* Equation \@ref(eq:random-effect-model) into Equation \@ref(eq:meta-regression). Now the product between $\beta_{1}X_{1}$ will consider the *lab vs. online effect*. To simulate this meta-regression, we need to fix the $\beta_{1}$ effect. Due to the fact that we are on a standardized scale, we can fix $\beta_{1}$ to be the expected difference between the two groups of studies. Crucially, despite $\tau^{2}$ is still the heterogeneity between effect sizes, now we need to fix $\tau^{2}$ considering that we included a moderator. In other terms, $\theta_i \sim \mathcal{N}(0, \tau_{r}^{2})$ where $\tau^2_r$ is the residual heterogeneity after including the moderator. We can describe our model using equation \@ref(eq:meta-regression-equations) according to the value of $X_{1i}$ (L = lab-based experiments, O = online experiments).

@@

\begin{align}
\begin{gathered}
y_{iL} = \beta_0 + \theta_i + \beta_1 \times 0 + \epsilon_i \\
d_{iO} = \beta_0 + \theta_i + \beta_1 \times 1 + \epsilon_i \\
(\#eq:meta-regression-equations)
\end{gathered}
\end{align}

@@

sim$b0_theta_i_x <- b0 + sim$theta_i + sim$b1*sim$age0

sim <- sim_studies(sim$b0_theta_i_x, sim$nc, sim$nt, data = sim)
```

@@

sim$theta_i <- rnorm(k, 0, sqrt(tau2r))
sim$age0 <- scale(sim$age, center = TRUE, scale = TRUE) # standardize the moderator
sim$b0_theta_i_x <- b0 + sim$theta_i + b1*sim$age0

sim <- sim_studies(sim$b0_theta_i_x, sim$nt, sim$nc, data = sim)
res <- rma(yi, vi, mods = ~age0, method = "REML", data = sim)
```

@@

```{r power-analysis-example, cache = FALSE}
set.seed(seed)
nsim <- 5000 # number of simulations per condition (5000, higher is better)
k <- c(5, 15, 25, 35, 50) # number of studies
theta <- 0.3 # the average effect size
tau2 <- 0.3 # the heterogeneity
navg <- 20 # average sample size per study
nmin <- 10 # minimum sample size per study
alpha <- 0.05 # the alpha level

# creating all combinations
sim_grid <- tidyr::expand_grid(k, theta, tau2, navg, nmin, nsim)

@@

sim_grid_t <- sim_grid
names(sim_grid_t) <- c("k", "$\\theta$", "$\\tau^2$", "$n_{avg}$", "$n_{min}$", "nsim", "Power")

@@

```{r plot-power-analysis, echo = FALSE, fig.cap=get_caption()}
title <- latex2exp::TeX(sprintf("$n_{avg} = %.1f$, $\\tau^2 = %s$, $\\theta_r = %s$, $\\alpha = 0.05$", sim_grid$navg[1], tau2, theta_r, 0.05))

@@

(ref:img-fixed-vs-random) The difference between the assumptions of the *fixed* and *random-effects model*. Each distribution depicts the sampling distribution of $k = 4$ hypothetical studies ($i = 1$, $2$, ..., $4$) with a certain observed effect size $d_{i}$, sampling variability $\sigma_{i}^{2}$ and the 95% confidence interval (black segment). The *fixed-effects* plot on the left suggests that each observed effect size has the same underlying true effect $\theta_{f}$ (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be remarkably close to the real effect ($\theta_{f}$). The *random-effects model* on the right suggests that beyond the error term ($\epsilon_{i}$), each real effect size is composed of a fixed part (now $\theta_{r}$) and a random part ($\theta_{i}$, black segments below the sampling distributions). The random part is determined by $\tau^{2}$, which represents the variability of the effect size distribution at the population level. When $\tau^{2}$ is zero, the random-effects reduces to a fixed-effects model.

>>>>

```{r monte-carlo-simulation}
set.seed(123) # seed for simulation to ensure the reproducibility of results
es <- 0.5 # real mean difference in the populations
sigma <- 1 # real standard deviations in the population
n <- 30 # sample size for both groups
nsims <- 1e5 # number of simulations = 100000

# simulates the sampling distribution of the mean difference
d_mc <- replicate(nsims, expr = {
  g1 <- rnorm(n = n, mean = es, sd = sigma)
  g2 <- rnorm(n = n, mean = 0, sd = sigma)
  mean(g1) - mean(g2) # calculates the sample mean difference and returns the value
})

@@

```{r notation-tab, echo=FALSE, result = "asis"}
notation_table <- tibble::tribble(
     ~`Equation`,     ~`Code`,                                               ~`Description`,
       "$\\theta$",   "theta",                  "Fixed-effects model real effect size",
       "$\\mu$",      "mu",                        "Random-effects model real effect size",
        "$\\tau^2$",    "tau2",                            "Effect size heterogeneity",
       "$\\tau^2_r$",   "tau2r",                   "Residual effect size heterogeneity",
  "$\\bar y_{T, C}$",        "",               "Mean of the experimental/control group",
      "$s_{T, C}$",        "", "Standard deviation of the experimental/control group",
        "$n_{T, C}$",  "nt, nc",        "Sample size of the experimental/control group",
             "$d_i$",      "yi",                                 "Observed effect size",
      "$\\sigma_{\\epsilon_i}^{2}$",      "vi",                          "Observed sampling variance",
       "$\\theta_i$", "theta\\_i",              "Random effect for the study i",
        "$\\beta_0$",      "b0",                            "Meta-regression intercept",
         "$\\beta_1$",      "b1",                                "Meta-regression slope",
               "$k$",       "k",                                    "Number of studies",
     "$\\epsilon_i$",        "",             "Sampling error for the study i"
  )

@@

```{r sim-random-effect}
set.seed(seed)
k <- 30 # number of studies
n <- 20 # number of participants per group, per study
mu <- 0.3 # real effect size
tau2 <- 0.2 # the effect size heterogeneity

sim <- make_data(k = k, nc = n, nt = n, mu = mu)

@@

# adding the by-study adjustment to the average effect
sim$es <- sim$mu + sim$theta_i

@@

```{r random-effect-high-low-precision, echo = FALSE, fig.asp=0.8, fig.cap=get_caption()}
set.seed(seed)
k <- 10 # number of studies
mu <- 0.3 # real effect size
tau2 <- 0.2
n <- c(30, 500)

sim <- make_data(k = k, mu = mu)

@@

sim$es <- sim$mu + sim$theta_i
sim <- sim_studies(sim$es, sim$nc, sim$nt, data = sim)

@@

```{r sim-random-effect-i2}
set.seed(seed)
k <- 30 # number of studies
mu <- 0.3 # real effect size
n <- 30 # sample size per group, per study
I2 <- 0.6 # desired I2 value

v <- (n + n)/(n * n) + mu^2/(2 * (n + n - 2)) # typical within study variance
tau2 <- -((I2*v)/(I2 - 1))

sim <- make_data(k = k, nc = n, nt = n, mu = mu)

sim$theta_i <- rnorm(k, 0, sqrt(tau2))
sim$es <- sim$mu + sim$theta_i 
sim <- sim_studies(sim$es, sim$nc, sim$nt, data = sim)
res <- rma(yi, vi, method = "REML", data = sim)
```

@@

A common example of meta-regression is by including a categorical predictor, including information about study-level features. In our example, a group of studies uses an online memory task while others use a standard lab-based task. The Equation \@ref(eq:random-effect-model1) can be easily extended for a meta-regression model by including a variable encoding the type of task (online vs. lab-based) and the expected difference between the two levels of the moderator (i.e., the *lab vs. online effect*). In regression terms (see equation \@ref(eq:meta-regression)), we could use a dummy variable ($X_{1}$) that takes the value of 0 for the *lab-based task* (*L*) and a value of 1 for an *online task* (*O*)^[The *dummy coding* (also known as treatment coding) is the default in R and `metafor` but other coding schemes could be used [see @Schad2020-ht for an overview of contrast coding schemes]]. Now we fix $\beta_{1}$ to be the *lab vs. online effect* (i.e., the expected mean difference between the two groups of studies) and the product between $\beta_{1}X_{1}$ will consider the *lab vs. online effect*. Crucially, despite $\tau^{2}$ is still the heterogeneity between effect sizes, now we need to fix $\tau^{2}$ considering that we included a moderator. In other terms, $\theta_i \sim \mathcal{N}(0, \tau_{r}^{2})$ where $\tau^2_r$ is the residual heterogeneity after including the moderator. We can describe our model using equation \@ref(eq:meta-regression-equations) according to the value of $X_{1i}$ (L = lab-based experiments, O = online experiments).

@@

\begin{align}
\begin{gathered}
y_{iL} = \beta_0 + \theta_i + \beta_1 \times 0 + \epsilon_i \\
d_{iO} = \beta_0 + \theta_i + \beta_1 \times 1 + \epsilon_i
(\#eq:meta-regression-equations)
\end{gathered}
\end{align}

@@

sim$es <- b0 + sim$theta_i + sim$b1*sim$age0

sim <- sim_studies(sim$es, sim$nc, sim$nt, data = sim)
```

@@

sim$theta_i <- rnorm(k, 0, sqrt(tau2r))
sim$age0 <- scale(sim$age, center = TRUE, scale = TRUE) # standardize the moderator
sim$es <- b0 + sim$theta_i + b1*sim$age0

sim <- sim_studies(sim$es, sim$nt, sim$nc, data = sim)
res <- rma(yi, vi, mods = ~age0, method = "REML", data = sim)
```

@@

```{r power-analysis-example, cache = TRUE}
set.seed(seed)
nsim <- 5000 # number of simulations per condition (5000, higher is better)
k <- c(5, 15, 25, 35, 50) # number of studies
delta <- 0.3 # the average effect size
tau2 <- 0.3 # the heterogeneity
navg <- 20 # average sample size per study
nmin <- 10 # minimum sample size per study
alpha <- 0.05 # the alpha level

# creating all combinations
sim_grid <- tidyr::expand_grid(k, mu, tau2, navg, nmin, nsim)

@@

sim_grid_t <- sim_grid
names(sim_grid_t) <- c("k", "$\\Delta$", "$\\tau^2$", "$n_{avg}$", "$n_{min}$", "nsim", "Power")

@@

```{r plot-power-analysis, echo = FALSE, fig.cap=get_caption()}
title <- latex2exp::TeX(sprintf("$n_{avg} = %.1f$, $\\tau^2 = %s$, $\\Delta = %s$, $\\alpha = 0.05$", sim_grid$navg[1], tau2, delta, 0.05))

@@

(ref:img-fixed-vs-random) The difference between the assumptions of the *fixed* and *random-effects model*. Each distribution depicts the sampling distribution of $k = 4$ hypothetical studies ($i = 1$, $2$, ..., $4$) with a certain observed effect size $y_{i}$ (red squares), sampling variability $\sigma_{i}^{2}$ and the 95% confidence interval (black segment). The *fixed-effects* plot on the left suggests that each observed effect size has the same underlying true effect $\theta$ (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be remarkably close to the real effect ($\theta$). The *random-effects model* on the right suggests that beyond the error term ($\epsilon_{i}$), each real effect size is composed of a fixed part (now $\mu_{\theta}$) and a random part ($\theta_{i}$, black segments below the sampling distributions). The random part is determined by $\tau^{2}$, which represents the variability of the effect size distribution at the population level. When $\tau^{2}$ is zero, the random-effects reduces to a fixed-effects model.
